A forest is more than its trees: haplotypes and inferred ARGs

### 2024-11-28 paper/writeup.tex
citations

There has been substantial recent progress
in the problem of ``ARG inference'',
which seeks to infer (portions of) the ``ancestral recombination graph'' (or, ARG)
that describes how a set of genotypes samples are related to each other
at each position of the genome \citep{speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust,lewanski2024introduction}.
One way of viewing the ARG is as a sequence of ``marginal trees'',
i.e., the genealogical trees that describe how each portion of the genome
was inherited by the focal genomes.
This is reflected in methodology of some ARG inference methods \citep[e.g., Relate][]{speidel2019method},
in metrics used to assess inference accuracy \citep{XXX},
as well as in basic terminology.
For instance, 
the ``succinct tree sequence'',
introduced by \citet{kelleher2016efficient},
is a common format for describing these inferred ARGs,
and is seeing wide use thanks in part to its efficiency and accompanying reliable toolkit,
\tskit \citep{XXX}.

@@

To do this, we simulate ARGs containing full haplotypes using \msprime's 
option \texttt{coalescing\_segments\_only} set to False.
Although \msprime{} simulates many events that do not create a coalescence in some marginal tree,
it only outputs information for nodes which contain a coalescence
(i.e., are the MRCA of some pair of samples at some point on the genome).
Furthermore, by default it only outputs those segments of the genome
on which there is a coalescence.
Said another way, by default all ancestral nodes in the ARG
output by \msprime{} are the MRCA of some pair of samples at every point in the genome
on which they are represented.
However, here we are interested in those segments of genome
on which the nodes are \emph{not} coalescent;
i.e., where they are unary in the marginal trees.
More concretely, we hope to recover those portions of ancestral haplotypes
on which the nodes are unary, but adjacent to a region of the genome where the node is not unary.
For instance, if a lineage carrying an ancestral segment of genome that spans $[a, c)$
coalesces with another spanning $[b, d)$, with $a < b < c < d$,
then the resulting node is only coalescent on $[b, c)$ but we hope using this algorithm
to extend the node's span to $[a, b)$ and $[c, d)$
(on which the node is unary).
However, following this example, the first lineage might carry a segment $[x, y)$
that is disjoint from the segment $[a, d)$.
We call these segments ``isolated non-coalescent segments'';
they have also been called ``trapped unary spans'' \citep[by][]{XXX}.
Such isolated segments will not be recovered by our algorithm,
and would likely be unrecoverable by any other method.
So, when evaluating what proportion of the true ARG is represented in an inferred ARG
we omit these isolated non-coalescent segments.
To give an idea of what proportion of the full spans of ancestral nodes
these isolated non-coalescent segments represent,
a simulation of 1000 samples
with genome length $5\times 10^7$, recombination rate $10^{-8}$, and population size $10^4$
has about half the total span of all nodes in isolated, non-coalescent segments.
For more discussion of these segments, see \citet{baumdicker2021efficient}.

We used simulations of several scenarios.
To include the effects of heterogeneous recombination rate,
in some we used stdpopsim \citep{stdpopsim} 
to simulate chromosome 1 of \textit{Canis familiaris}
using the CanFam3 genetic map from \citet{campell2016}.
\emph{``Constant dog''} simulations simulated this chromosome in a 
population of (constant) size $10^4$.
\emph{``Expanding dog''} simulations were similar, but
used a discrete-time Wright--Fisher model to simulate
a small population of 100 that expanded to 1,000 individuals ten generations ago,
which then doubled every generation to reach 512,000 individuals.
All jobs for which runtime was recorded were executed on an Intel Xeon Gold 6148 processor.
 Additionally, we compare accuracy between sequences modified from a ``true'' tree sequence
 using our dissimilarity methods. 
 These tree sequences have a population of 1,000 and recombination rate of $10^{-8}$ with samples ranging from 10 to 1,000 and genome length ranging from $10^6$ to $5\times 10^7$.

>>>>

There has been substantial recent progress
in the problem of ``ARG inference'',
which seeks to infer (portions of) the ``ancestral recombination graph'' (or, ARG)
that describes how a set of genotypes samples are related to each other
at each position of the genome \citep{speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust,lewanski2024introduction}.
One way of viewing the ARG is as a sequence of ``marginal trees'',
i.e., the genealogical trees that describe how each portion of the genome
was inherited by the focal genomes.
This is reflected in methodology of some ARG inference methods
and in metrics used to assess inference accuracy,
as well as in basic terminology.
For instance, 
the ``succinct tree sequence'',
introduced by \citet{kelleher2016efficient},
is a common format for describing these inferred ARGs,
and is seeing wide use thanks in part to its efficiency and accompanying reliable toolkit,
\tskit{} \citep{tskit}.

@@

To do this, we simulate ARGs containing full haplotypes using \msprime's 
option \texttt{coalescing\_segments\_only} set to False.
Although \msprime{} simulates many events that do not create a coalescence in some marginal tree,
it only outputs information for nodes which contain a coalescence
(i.e., are the MRCA of some pair of samples at some point on the genome).
Furthermore, by default it only outputs those segments of the genome
on which there is a coalescence.
Said another way, by default all ancestral nodes in the ARG
output by \msprime{} are the MRCA of some pair of samples at every point in the genome
on which they are represented.
However, here we are interested in those segments of genome
on which the nodes are \emph{not} coalescent;
i.e., where they are unary in the marginal trees.
More concretely, we hope to recover those portions of ancestral haplotypes
on which the nodes are unary, but adjacent to a region of the genome where the node is not unary.
For instance, if a lineage carrying an ancestral segment of genome that spans $[a, c)$
coalesces with another spanning $[b, d)$, with $a < b < c < d$,
then the resulting node is only coalescent on $[b, c)$ but we hope using this algorithm
to extend the node's span to $[a, b)$ and $[c, d)$
(on which the node is unary).
However, following this example, the first lineage might carry a segment $[x, y)$
that is disjoint from the segment $[a, d)$.
We call these segments ``isolated non-coalescent segments'';
they have also been called ``trapped unary spans'' \citep[by][]{wong2024general}.
Such isolated segments will not be recovered by our algorithm,
and would likely be unrecoverable by any other method.
So, when evaluating what proportion of the true ARG is represented in an inferred ARG
we omit these isolated non-coalescent segments.
To give an idea of what proportion of the full spans of ancestral nodes
these isolated non-coalescent segments represent,
a simulation of 1000 samples
with genome length $5\times 10^7$, recombination rate $10^{-8}$, and population size $10^4$
has about half the total span of all nodes in isolated, non-coalescent segments.
For more discussion of these segments, see \citet{baumdicker2021efficient}.

We used simulations of several scenarios.
To include the effects of heterogeneous recombination rate,
in some we used stdpopsim \citep{adrion2020stdpopsim} 
to simulate chromosome 1 of \textit{Canis familiaris}
using the CanFam3 genetic map from \citet{campbell2016}.
\emph{``Constant dog''} simulations simulated this chromosome in a 
population of (constant) size $10^4$.
\emph{``Expanding dog''} simulations were similar, but
used a discrete-time Wright--Fisher model to simulate
a small population of 100 that expanded to 1,000 individuals ten generations ago,
which then doubled every generation to reach 512,000 individuals.
All jobs for which runtime was recorded were executed on an Intel Xeon Gold 6148 processor.
 Additionally, we compare accuracy between sequences modified from a ``true'' tree sequence
 using our dissimilarity methods. 
 These tree sequences have a population of 1,000 and recombination rate of $10^{-8}$ with samples ranging from 10 to 1,000 and genome length ranging from $10^6$ to $5\times 10^7$.
